---
title: "Phenological Shifts in the Ruby-Throated Hummingbird and Preferred Flowers"
author: Aubrey Monaco
subtitle: GEO511 Final Project
date: today
date-format: long
editor_options:
  chunk_output_type: console
---

# Introduction

With warming temperatures due to anthropogenic climate change, bird migration times are shifting, and they are tending to migrate back to their breeding grounds earlier each year. Conversely, flowers that certain migrating birds, like Ruby-throated Hummingbirds (Archilochus colubris), feed on may not be changing the time of their first bloom. Therefore, when birds that rely on flowers as food sources migrate back earlier, they may not have enough food to support them when they arrive. I want to investigate how a pollinating bird species, the Ruby-throated Hummingbird, has been shifting its migration patterns and how the appearances of species it prefers to feed on, Red Buckeye (Aesculus pavia) and columbine (Aquilegia canadensis), has changed (or not changed) over the past 8 years. I predict that Ruby-throated Hummingbirds are migrating back earlier each year, and they are arriving at increasing latitudes earlier than lower latitudes. Additionally, I suppose that the first sightings of these flowers in the months that Ruby-throated Hummingbirds migrate back North will be earlier compared to previous years, but not to the same extent as the bird migrations.


# Materials and methods

Narrative: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.

1.  I loaded my necessary libraries

```{r, message=FALSE, warning=FALSE}
#install.packages("leaflet")
#install.packages("ebirdst")
#install.packages("rinat")
#install.packages("rnpn")
library(leaflet)
library(ggplot2)
library(dplyr)
library(sf)
library(spData)
library(lubridate)
library(ebirdst)
library(rinat)
library(rnpn)

```

2. I loaded my data from iNaturalist and the US National Phenology Network and joined my data into one set.

```{r, message=FALSE, warning=FALSE}
#rb2010 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2010, place_id = 1, maxresults = 5000, quality = "research")
#cb2010 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2010, place_id = 1, maxresults = 5000, quality = "research")

#fl_2010 <- rbind(rb2010, cb2010)

#rb2011 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2011, place_id = 1, maxresults = 5000, quality = "research")
#cb2011 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2011, place_id = 1, maxresults = 5000, quality = "research")

#fl_2011 <- rbind(rb2011, cb2011)

#rb2012 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2012, place_id = 1, maxresults = 5000, quality = "research")
#cb2012 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2012, place_id = 1, maxresults = 5000, quality = "research")

#fl_2012 <- rbind(rb2012, cb2012)

#rb2013 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2013, place_id = 1, maxresults = 5000, quality = "research")
#cb2013 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2013, place_id = 1, maxresults = 5000, quality = "research")

#fl_2013 <- rbind(rb2013, cb2013)

#rb2014 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2014, place_id = 1, maxresults = 5000, quality = "research")
#cb2014 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2014, place_id = 1, maxresults = 5000, quality = "research")

#fl_2014 <- rbind(rb2014, cb2014)

#rb2015 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2015, place_id = 1, maxresults = 5000, quality = "research")
#cb2015 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2015, place_id = 1, maxresults = 5000, quality = "research")

#fl_2015 <- rbind(rb2015, cb2015)

#rb2016 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2016, place_id = 1, maxresults = 5000, quality = "research")
#cb2016 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2016, place_id = 1, maxresults = 5000, quality = "research")

#fl_2016 <- rbind(rb2016, cb2016)

#rb2017 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2017, place_id = 1, maxresults = 5000, quality = "research")
#cb2017 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2017, place_id = 1, maxresults = 5000, quality = "research")

#fl_2017 <- rbind(rb2017, cb2017)

#rb2018 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2018, place_id = 1, maxresults = 5000, quality = "research")
#cb2018 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2018, place_id = 1, maxresults = 5000, quality = "research")

#fl_2018 <- rbind(rb2018, cb2018)

#rb2019 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2019, place_id = 1, maxresults = 5000, quality = "research")
#cb2019 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2019, place_id = 1, maxresults = 5000, quality = "research")

#fl_2019 <- rbind(rb2019, cb2019)

#rb2020 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2020, place_id = 1, maxresults = 5000, quality = "research")
#cb2020 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2020, place_id = 1, maxresults = 5000, quality = "research")

#fl_2020 <- rbind(rb2020, cb2020)

#rb2021 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2021, place_id = 1, maxresults = 5000, quality = "research")
#cb2021 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2021, place_id = 1, maxresults = 5000, quality = "research")

#fl_2021 <- rbind(rb2021, cb2021)

#rb2022 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2022, place_id = 1, maxresults = 5000, quality = "research")
#cb2022 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2022, place_id = 1, maxresults = 5000, quality = "research")

#fl_2022 <- rbind(rb2022, cb2022)

#rb2023 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2023, place_id = 1, maxresults = 5000, quality = "research")
#cb2023 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2023, place_id = 1, maxresults = 5000, quality = "research")

#fl_2023 <- rbind(rb2023, cb2023)

#rb2024 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2024, place_id = 1, maxresults = 5000, quality = "research")
#cb2024 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2024, place_id = 1, maxresults = 5000, quality = "research")

#fl_2024 <- rbind(rb2024, cb2024)

#rb2025 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2025, place_id = 1, maxresults = 5000, quality = "research")
#cb2025 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2025, place_id = 1, maxresults = 5000, quality = "research")

#fl_2025 <- rbind(rb2025, cb2025)

#fl_all <- rbind(fl_2010, fl_2011, fl_2012, fl_2013, fl_2014, fl_2015, fl_2016, fl_2017, fl_2018, fl_2019, fl_2020, fl_2021, fl_2022, fl_2023, fl_2024, fl_2025)

#write.csv(fl_all, "data/fl_all.csv")

nat_flower <- read.csv("data/fl_all.csv")
nat_flower <- nat_flower %>%
  filter(captive_cultivated == "false")

#Species IDs from USNPN

#npn_download_individual_phenometrics(
 #  request_source='Aubrey Monaco', 
 # years=c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020,2021,2022,2023,2024, 2025), 
#  species_id=c(784, 149), 
#  download_path="npn_flower.csv"
# )

npn_flower <- read.csv("npn_flower.csv")

all_flower <- nat_flower %>% full_join(npn_flower)

```

3. I used the Lubridate package to standardize the format all of the dates in my flower data. Then, I used this standardized data to create a year column, where I pulled the years from the dates, and a year_day column, to transform the date into a day out of 365, and a month column to pull the month from the dates.
```{r, message=FALSE, warning=FALSE}
all_flower <- all_flower %>%
  mutate(observed_on = parse_date_time(observed_on, orders = c("mdy", "ydm", "dmy"))) %>%
  mutate(year = year(observed_on)) %>%
  mutate(year_day = yday(observed_on)) %>%
  mutate(month = month(observed_on))
 
```

4. I found the mean date of first sightings of the selected flowers from 2010 to 2017 and from June to September (when ruby-throats are most abundant) to compare data from the last 8 years to. 

```{r, message=FALSE, warning=FALSE}
fl_early <- all_flower %>%
  filter(year %in% c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017) & month %in% c(2, 3, 4, 5, 6))

fl_early_27 <- fl_early %>%
  filter(latitude > 27 & latitude < 33) %>%
  mutate(mean27 = mean(year_day, na.rm = TRUE)) 

fl_early_33 <- fl_early %>%
  filter(latitude > 33 & latitude < 36) %>%
  mutate(mean33 = mean(year_day, na.rm = TRUE))

fl_early_36 <- fl_early %>%
  filter(latitude > 36 & latitude < 39) %>%
  mutate(mean36 = mean(year_day, na.rm = TRUE))

fl_early_39 <- fl_early %>%
  filter(latitude > 39 & latitude <= 42) %>%
  mutate(mean39 = mean(year_day, na.rm = TRUE)) 

fl_early_42 <- fl_early %>%
  filter(latitude > 42 & latitude < 59) %>%
  mutate(mean42 = mean(year_day, na.rm = TRUE))

flearly1 <- fl_early_27 %>% full_join(fl_early_33)
flearly2 <- fl_early_36 %>% full_join(fl_early_39)
flearly3 <- flearly1 %>% full_join(fl_early_42)

flearly <- flearly2 %>% full_join(flearly3)

flower_early_all <- all_flower %>% full_join(flearly)
```


5. I created the dataset to see trends between 2018 and 2025, and in months Janurary to May. I separated sightings into 5 bands of latitude (in increments of approxiamtely 3, excepting 27 to 33, since there were so few observations). I rejoined all of these bands into one dataset to plot on a map. I also created a difference column, showing the difference in days of arrival between the new dataset and the 2010 to 2017 average using mean - mean1017.

```{r, message=FALSE, warning=FALSE}
flower_new <- flower_early_all %>%
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & month %in% c(2, 3, 4, 5, 6)) %>%
  mutate(mean1017 = case_when(latitude >= 27 & latitude < 33 ~ 81.54737
,
                          latitude > 33 & latitude < 36 ~ 85.73214,
                          latitude > 36 & latitude < 39 ~ 88.62222,
                          latitude > 39 & latitude < 42 ~ 92.59701,
                          latitude > 42 & latitude < 59 ~ 94.25773))

fl_27 <- flower_new %>%
  filter(latitude > 27 & latitude < 33) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

fl_33 <- flower_new %>%
  filter(latitude > 33 & latitude < 36) %>%
  mutate(av = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = av - mean1017)

fl_36 <- flower_new %>%
  filter(latitude > 36 & latitude < 39) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

fl_39 <- flower_new %>%
  filter(latitude > 39 & latitude < 42) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

fl_42 <- flower_new %>%
  filter(latitude > 42 & latitude < 50) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

fl_1 <- fl_27 %>% full_join(fl_33)
fl_2 <- fl_36 %>% full_join(fl_39)
fl_3 <- fl_42 %>% full_join(fl_2)

fl_4 <- fl_1 %>% full_join(fl_2)

fl_all <- fl_3 %>%
  full_join(fl_4)

```

6. I followed the same steps to find data for ruby-throated hummingbird migration and found their average arrival dates for various latitudes between 2010 and 2017. I used data from eBird in addition to iNaturalist and the US National Phenology Network.

My eBird datasets can be found on figshare here:
- 1995-2021: https://figshare.com/articles/dataset/eBird_rthhum_2010_2014/30535889?file=59303393
- 2022-2025: https://figshare.com/articles/dataset/eBird_rthhum_2015_2025/30535895?file=59303390

```{r, message=FALSE, warning=FALSE}

#rth2010 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2010, place_id = 1, maxresults = 5000)

#rth2011 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2011, place_id = 1, maxresults = 5000)

#rth2012 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2012, place_id = 1, maxresults = 5000)

#rth2013 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2013, place_id = 1, maxresults = 5000)

#rth2014 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2014, place_id = 1, maxresults = 5000)

#rth2015 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2015, place_id = 1, maxresults = 5000)

#rth2016 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2016, place_id = 1, maxresults = 5000)

#rth2017 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2017, place_id = 1, maxresults = 5000)

#rth2018 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2018, place_id = 1, maxresults = 5000)

#rth2019 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2019, place_id = 1, maxresults = 5000)

#rth2020 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2020, place_id = 1, maxresults = 5000)

#rth2021 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2021, place_id = 1, maxresults = 5000)

#rth2022 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2022, place_id = 1, maxresults = 5000)

#rth2023 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2023, place_id = 1, maxresults = 5000)

#rth2024 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2024, place_id = 1, maxresults = 5000)

#rth2025 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2025, place_id = 1, maxresults = 5000)

#rth_nat <- rbind(rth2010, rth2011, rth2012, rth2013, rth2014, rth2015, rth2016, rth2017, rth2018, rth2019, rth2020, rth2021, rth2022, rth2023, rth2024, rth2025)

#write.csv(rth_nat, "data/rth_nat.csv")

#npn_download_individual_phenometrics(
# request_source='Aubrey Monaco', 
# years=c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020,2021,2022,2023,2024, 2025), 
#species_id=c(352), 
# download_path="data/npn_rth.csv")

npn_rth <- read.csv("data/npn_rth.csv")
npn_rth <- npn_rth %>%
  rename(Latitude = latitude)

rth_1014 <- read.csv("data/RTH_1995_2021.csv")
rth_1525 <- read.csv("data/RTH_2022_2025.csv")

rth_ebird <- rth_1014 %>% full_join(rth_1525)

rth_inat <- read.csv("data/rth_nat.csv")
rth_inat <- rth_inat %>%
  rename(Latitude = latitude) %>%
  filter(captive_cultivated == "false" & quality_grade == "research")

rth_some <- rth_ebird %>% full_join(rth_inat)

rth_all <- rth_some %>% full_join(npn_rth)

ruby_throat <- rth_all %>%
  mutate(date_new = parse_date_time(Date, orders = c("mdy", "ydm", "dmy"))) %>%
  mutate(year_new = year(date_new)) %>%
  mutate(year_day = yday(date_new))

rth_early <- ruby_throat %>%
  filter(Year %in% c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017) & Month %in% c(2, 3, 4, 5))

rthearly30 <- rth_early %>%
  filter(Latitude > 30 & Latitude < 33) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rthearly33 <- rth_early %>%
  filter(Latitude > 33 & Latitude < 36) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rthearly36 <- rth_early %>%
  filter(Latitude > 36 & Latitude < 39) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rthearly39 <- rth_early %>%
  filter(Latitude > 39 & Latitude < 42) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 
  
rthearly42 <- rth_early %>%
  filter(Latitude > 42 & Latitude < 46) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 
  

rthearly1 <- rthearly30 %>% full_join(rthearly33)
rthearly2 <- rthearly36 %>% full_join(rthearly39)
rthearly3 <- rthearly42 %>% full_join(rthearly2)

rthearly <- rthearly1 %>% full_join(rthearly3)

ruby_throat_all <- ruby_throat %>% full_join(rth_early)

```

7. I then similarly created the runy-throat dataset for 2018-2025, but I added 3 more bands of latitude to account for the larger number of observations recorded at higher and lower latitudes. I again combined these bands into one dataset, and created a difference column, showing the difference in days of arrival between the new dataset and the 2010 to 2017 average using mean - mean1017.

```{r, message=FALSE, warning=FALSE}

rth_all <- ruby_throat_all %>%
  filter(Year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & Month %in% c(2, 3, 4, 5)) %>%
  mutate(mean1017 = case_when(
                              Latitude > 30 & Latitude < 33 ~ 99.69231,
                              Latitude > 33 & Latitude < 36 ~ 104.48,
                              Latitude > 36 & Latitude < 39 ~ 125.4231,
                              Latitude > 39 & Latitude < 42 ~ 133.2071,
                              Latitude > 42 & Latitude < 49 ~ 138.1325
                              ))

rth_30 <- rth_all %>%
  filter(Latitude > 30 & Latitude < 36) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

rth_33 <- rth_all %>%
  filter(Latitude > 33 & Latitude < 36) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

rth_36 <- rth_all %>%
  filter(Latitude > 36 & Latitude < 39) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

rth_39 <- rth_all %>%
  filter(Latitude > 39 & Latitude < 42) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

rth_42 <- rth_all %>%
  filter(Latitude > 42 & Latitude < 45) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean1017)

rth_1 <- rth_30 %>% full_join(rth_33)
rth_2 <- rth_36 %>% full_join(rth_39)
rth_3 <- rth_2 %>% full_join(rth_42)

rubythroat_all <- rth_1 %>% full_join(rth_3)

```

8. I then created color palettes using viridis scales and colorNumeric from the Leaflet package for my two maps, specifying to use the difference column in each dataset to map the colors to.

```{r}

cpal <- colorNumeric(palette = "mako", domain = fl_all$difference, na.color = NA)

rpal <- colorNumeric(palette = "mako", domain = rubythroat_all$difference, na.color = NA)
 
```


### Trends in Flower Blooms (2018 - 2025)
```{r}
leaflet(fl_all) %>%
  addProviderTiles(provider = "Esri") %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~longitude, ~ latitude, radius = 1, color = ~cpal(difference)) %>%
  addLegend(pal = cpal, values = ~difference, title = "Difference From 2010-2017 Average (days)", position = "bottomleft")
```


### Trends in Ruby-Throated Hummingbird Sightings (2018 - 2025)
```{r}
leaflet(rubythroat_all) %>%
  addProviderTiles(provider = "Esri") %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~Longitude, ~ Latitude, radius = 1, color = ~rpal(difference)) %>%
  addLegend(pal = rpal, values = ~difference, title = "Difference From 2010-2017 Average (days)", position = "bottomleft") 
```


## Download and clean all required data

To download Ruby-throated Hummingbird data from eBird on your own, visit: https://media.ebird.org/catalog?birdOnly=true&taxonCode=rthhum, filter by year (I did this in two separate actions [1995-2021 and 2022-2025] to ensure I could download all of the data), and filter by location (United States)

# Results

\[\~200 words\]

Tables and figures (maps and other graphics) are carefully planned to convey the results of your analysis. Intense exploration and evidence of many trials and failures. The author looked at the data in many different ways before coming to the final presentation of the data.

Trends for the preferred flowers have shown that its timing of first observations have changed from the 2010-2014 average, and at most latitudes, when moving higher, there is a pattern of earlier sightings each year. But for the lowest latitude, flowers have been sighted earlier than some higher latitudes. This may be due to hummingbirds 

Show tables, plots, etc. and describe them.r

# Conclusions

\[\~200 words\]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner

Thanks for checking out my web site!
