---
title: "Phenological Changes in Pollination Pals"
author: Aubrey Monaco
subtitle: Synchrony Shifts in the Ruby-Throated Hummingbird and Preferred Flowers
date: today
date-format: long
editor_options:
  chunk_output_type: console
---

# Introduction

With warming temperatures due to anthropogenic climate change, migrating birds are tending to move back to their breeding grounds earlier each year [1]. Along with this, flowers that certain migrating birds, like Ruby-throated Hummingbirds (Archilochus colubris), feed on may be changing the time of their first bloom due to climate change [2].

![[Ruby-throated Hummingbird [3]]{style="color:darkblue;"}](data/rth_1.jpg){fig-align="center" width="400"}

I want to investigate how the Ruby-throated Hummingbird has been shifting its spring migration patterns, and how the appearances of two spring ephemeral species it prefers to feed on during this migration, Red Buckeye (Aesculus pavia) and Red Columbine (Aquilegia canadensis), has changed (or not changed) over the past 10* years [4].

*My maps and graphs are from 2014 to 2025 to account for any incomplete datasets from 2025

![[Red Columbine [5]]{style="color:darkred;"}](data/cb_1.jpg){fig-align="center" width="200"}

![[Red Buckeye [6]]{style="color:darkred;"}](data/rb_1.jpg){fig-align="center" width="226"}

I predict that Ruby-throated Hummingbirds are migrating back to their breeding grounds earlier each year, and that they are arriving at higher latitudes increasingly earlier than lower latitudes. Additionally, I suppose that the first sightings of these preferred flowers will be earlier at higher latitudes and later at lower latitudes compared to previous years. So, as Ruby-throated Hummingbirds migrate back earlier, these flowers may not be available at lower latitudes for them to feed on.

# Materials and methods

## 1. Install Packages and Load Libraries

```{r, message=FALSE, warning=FALSE}

#install.packages("leaflet") # Dynamic maps
#install.packages("leaflet.providers") # Different Basemaps
#install.packages("leaflet.extras") # Heatmaps
#install.packages("rinat") # Downloads iNaturalist data
#install.packages("rnpn") # Downloads US National Phenology Network Data
#install.packages("plotly") # Interactive graphs
library(doParallel) # Download data in a parallel loop
library(foreach) # Download data in a lpp[]
library(plotly)
library(leaflet)
library(leaflet.providers)
library(leaflet.extras)
library(rinat)
library(rnpn)
library(sf)
library(rvest) # Used to download data from Journey North
library(dplyr) # Mutating and filtering data
library(lubridate) # Parsing date data
library(ggplot2) # Plotting trends

```

## 2. Flowers: Load iNaturalist and US NPN Data.

- foreach, doParallel, and rinat package iNaturalist data download: 

- buck = Red Buckeye

- col = Red Columbine

- taxon_name is set to each desired species, place_id is set to 1 for the United States, maxresults is set to 10000 to ensure all data is downloaded, and quality is set to "research" to ensure verified data

*The functions are commented out after running them once and writing the data into a csv, so that they do not start each time the code is run

```{r message=FALSE, warning=FALSE, eval=FALSE}
#| code-fold: TRUE

#Registering cores for downloading data

num_cores <- parallel::detectCores() - 1
registerDoParallel(num_cores)

#Data for both species 2005 to 2018

nat_0518 <- foreach(s = c("Aesculus pavia", "Aquilegia canadensis"), .packages = c("dplyr", "rinat"), .errorhandling = "pass", .combine = bind_rows) %:% foreach(a = c(2005:2018), .packages = c("dplyr", "rinat"), .errorhandling = "pass", .combine = bind_rows) %do% { 
print(paste("Downloading data for year:", a))
Sys.sleep(45)
get_inat_obs(taxon_name = s, year = a, place_id = 1, maxresults = 10000, quality = "research", query = n) 
}

#Red Buckeye Data 2019, 2020

buck_1920 <- foreach(b = c(2019,2020), .packages = c("dplyr", "rinat"), .combine = bind_rows, .errorhandling = "pass") %dopar% {
print(paste("Downloading data for year:", b))
Sys.sleep(30)
get_inat_obs(taxon_name = "Aesculus pavia", year = b, place_id = 1, maxresults = 10000, quality = "research") 
}

#Red Buckeye Data 2021, 2022

buck_2122 <- foreach(c = c(2021,2022), .packages = c("dplyr", "rinat"), .combine = bind_rows) %dopar% {
print(paste("Downloading data for year:", c))
Sys.sleep(30)
get_inat_obs(taxon_name = "Aesculus pavia", year = c, place_id = 1, maxresults = 10000, quality = "research") 
}

#Red Buckeye Data 2023, 2024

buck_2324 <- foreach(c = c(2023,2024), .packages = c("dplyr", "rinat"), .combine = bind_rows) %dopar% {
print(paste("Downloading data for year:", c))
Sys.sleep(30)
get_inat_obs(taxon_name = "Aesculus pavia", year = c, place_id = 1, maxresults = 10000, quality = "research") 
}

#Red Columbine Data 2019, 2020

col_1920 <- foreach(d = c(2019,2020), .packages = c("dplyr", "rinat"), .combine = bind_rows, .errorhandling = "pass") %dopar% {
print(paste("Downloading data for year:", d))
Sys.sleep(30)
get_inat_obs(taxon_name = "Aquilegia canadensis", year = d, place_id = 1, maxresults = 10000, quality = "research") 
}

#Red Columbine Data 2021-25

col_21 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2021, place_id = 1, maxresults = 10000, quality = "research")
col_22 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2022, place_id = 1, maxresults = 10000, quality = "research") 
col_23 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2023, place_id = 1, maxresults = 10000, quality = "research") 
col_24 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2024, place_id = 1, maxresults = 10000, quality = "research") 
col_25 <- get_inat_obs(taxon_name = "Aquilegia canadensis", year = 2025, place_id = 1, maxresults = 10000, quality = "research")

#Red Buckeye Data 2025

buck_25 <- get_inat_obs(taxon_name = "Aesculus pavia", year = 2025, place_id = 1, maxresults = 10000, quality = "research")

```

- Individual datasets combined and written csv is loaded into environment as "nat_flower"

```{r, message=FALSE, warning=FALSE, eval=FALSE}

fl_all <- rbind(nat_0518, buck_1920, buck_2122, buck_2324, col_1920, col_21, col_22, col_23, col_24, col_25, buck_25)

write.csv(fl_all, "data/fl_all.csv")

```

```{r}

nat_flower <- read.csv("data/fl_all.csv") 
nat_flower <- nat_flower %>%
  filter(common_name %in% c("red columbine", "Red Buckeye"))

```

- rnpn package for NPN data download (automatically saves as a csv): 

```{r, message=FALSE, warning=FALSE, eval=FALSE}

# Species IDs are available through the NPN's website (Red Buckeye = 784, Red Columbine = 147)

npn_download_individual_phenometrics(
 request_source='Aubrey Monaco', 
 years=c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
 species_id=c(784, 147), 
 download_path="data/npn_flower.csv"
 )

```

```{r}

npn_flower <- read.csv("data/npn_flower.csv")

```

## 3.  Flowers: Combine the iNaturalist and NPN Data

```{r}

# iNaturalist and NPN data are joined and duplicate observations are removed

all_flower <- nat_flower %>% full_join(npn_flower)
all_flower <- all_flower[!duplicated(as.list(all_flower))]
all_flower <- all_flower %>%
  distinct()

```

## 4. Flowers: Standardize and Mutate Data

- dplyr and lubridate used to standardize and add columns from dates

```{r, message=FALSE, warning=FALSE}

all_flower <- all_flower %>%
  mutate(observed_on = parse_date_time(observed_on, orders = c("mdy", "ydm", "dmy"))) %>% # Standardizes date formats
  mutate(year = year(observed_on)) %>% # Pulls year from date
  mutate(year_day = yday(observed_on)) %>% # Puts date into day out of 365
  mutate(month = month(observed_on)) # Pulls month from date
 
```

## 5. Flowers: Calculate Means of Early Period (2005-2013)

- Found the mean date of first flower sightings from 2005 to 2013 and from March to May (when these flowers' bloom times overlap) [7,8]
- Separated sightings into 3 bands of latitude (grouped latitudes had similar averages)

```{r, message=FALSE, warning=FALSE}

fl_early <- all_flower %>%
  filter(year %in% c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013) & month %in% c(3, 4, 5))

fl_early_all <- fl_early %>%
  mutate(latbin=cut(latitude, c(30,36,42,48))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE))

fl_means <- fl_early_all %>%
  group_by(latbin) %>%
  summarise(mean)

```

## 6. Flowers: Calculate Means and Difference From 2014-2025 to 2005-2013

- Added a column (mean0013) with the means of the fl_early observations based on latitude
- Used 6 bands of latitude for more specific analysis
- Created a "difference" column, showing the difference in days of arrival between the early and current average using mean - mean0013
- Binded all of these bands into one dataset for plotting

```{r, message=FALSE, warning=FALSE}
fl_new <- all_flower %>%
  filter(year %in% c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & month %in% c(3, 4, 5)) %>%
  mutate(mean0013 = 
         case_when(
         latitude > 30 & latitude < 33 ~ 81.95652,
         latitude > 33 & latitude < 36 ~ 81.95652,
         latitude > 36 & latitude < 39 ~ 99.73333,
         latitude > 39 & latitude < 42 ~ 99.73333,
         latitude > 42 & latitude < 45 ~ 114.30769,
         latitude > 45 & latitude < 48 ~ 114.30769))

fl_new_all <- fl_new %>%
  mutate(latbin=cut(latitude, c(30,33,36,39,42,45,48))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE)) %>%
  mutate(difference = mean-mean0013)

```

## 7. Repeat Process for Ruby-throated Hummingbirds

- Used data from eBird and Journey North in addition to iNaturalist and the US NPN.

My eBird datasets can be found on figshare here:

-   1995-2021: <https://figshare.com/articles/dataset/eBird_rthhum_2010_2014/30535889?file=59303393>

-   2022-2025: <https://figshare.com/articles/dataset/eBird_rthhum_2015_2025/30535895?file=59303390>

Visit each link and click "download"

Or:

Download the Ruby-throated Hummingbird data from eBird by visiting: <https://media.ebird.org/catalog?birdOnly=true&taxonCode=rthhum>, filter by year (I did this in two separate actions \[1995-2021 and 2022-2025\] to ensure I could download all of the data), filter by location (United States), and download as a csv.

- iNaturalist data download:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
#| code-fold: TRUE

#Ruby-throat data 2005-2019

rth_0519 <- foreach(e = c(2005:2019), .packages = c("dplyr", "rinat"), .combine = bind_rows) %dopar% {
print(paste("Downloading data for year:", e))
Sys.sleep(45)
get_inat_obs(taxon_name = "Archilochus colubris", year = e, place_id = 1, maxresults = 10000, quality = "research") 
}

#Ruby-throat data 2020-2025

rth_20 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2020, place_id = 1, maxresults = 10000, quality = "research")
rth_21 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2021, place_id = 1, maxresults = 10000, quality = "research")
rth_22 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2022, place_id = 1, maxresults = 10000, quality = "research") 
rth_23 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2023, place_id = 1, maxresults = 10000, quality = "research") 
rth_24 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2024, place_id = 1, maxresults = 10000, quality = "research") 
rth_25 <- get_inat_obs(taxon_name = "Archilochus colubris", year = 2025, place_id = 1, maxresults = 10000, quality = "research") 

```

```{r, eval=FALSE}

#Combined all iNat data into one data set for Ruby-Throats

rth_nat <- rbind(rth_0519, rth_20, rth_21, rth_22, rth_23, rth_24, rth_25)

write.csv(rth_nat, "data/rth_nat.csv")

```

- NPN data download:

```{r, warning=FALSE, message=FALSE, eval=FALSE}
#| code-fold: TRUE

#Ruby-Throated Hummingbird = 352 (Species ID)

npn_download_individual_phenometrics(
 request_source='Aubrey Monaco', 
 years=c(2005, 2006, 2007, 2008, 2009, 2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024, 2025),
 species_id=c(352), 
 download_path="data/npn_rth.csv")

```

```{r}

#Renamed latitude column for easier manipulation

npn_rth <- read.csv("data/npn_rth.csv")
npn_rth <- npn_rth %>%
  rename(Latitude = latitude, Longitude = longitude)

```

Journey North Data Download:

```{r, collapse=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
#| code-fold: TRUE

#Links for all years of data download (can only do one at a time)

jn_05 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2005&submit=View+Data"
jn_06 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2006&submit=View+Data"
jn_07 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2007&submit=View+Data"
jn_08 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2008&submit=View+Data"
jn_09 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2009&submit=View+Data"
jn_10 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2010&submit=View+Data"
jn_11 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2011&submit=View+Data"
jn_12 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2005&submit=View+Data"
jn_13 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2013&submit=View+Data"
jn_14 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2014&submit=View+Data"
jn_15 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2015&submit=View+Data"
jn_16 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2016&submit=View+Data"
jn_17 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2017&submit=View+Data"
jn_18 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2018&submit=View+Data"
jn_19 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2019&submit=View+Data"
jn_20 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2020&submit=View+Data"
jn_21 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2021&submit=View+Data"
jn_22 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2022&submit=View+Data"
jn_23 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2023&submit=View+Data"
jn_24 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2024&submit=View+Data"
jn_25 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2025&submit=View+Data"

#Download data for all links

jn_table <- foreach(urls = c(jn_05, jn_06, jn_07, jn_08, jn_09, jn_10, jn_11, jn_12, jn_13, jn_14, jn_15, jn_16, jn_17, jn_18, jn_19, jn_20, jn_21, jn_22, jn_23, jn_24, jn_25), .packages = c("dplyr", "rvest"), .combine = bind_rows) %dopar% {
Sys.sleep(30)
read_html(urls) %>%
html_elements("table") %>%
html_table() %>%
as.data.frame()
}

```

- Joining all data:

```{r, warning=FALSE, message=FALSE}
#| code-fold: TRUE

#jn_table <- as.data.frame(jn_table)
#write.csv(jn_table, "data/jn_table.csv")
jn_table <- read.csv("data/jn_table.csv")

#Datasets from eBird
rth_0513 <- read.csv("data/RTH_1995_2021.csv")
rth_1425 <- read.csv("data/RTH_2022_2025.csv")

#Joined eBird data
rth_ebird <- rth_0513 %>% full_join(rth_1425)

#iNaturalist data loaded into environment and standardized
rth_inat <- read.csv("data/rth_nat.csv")
rth_inat <- rth_inat %>%
  rename(Latitude = latitude, Longitude = longitude) %>% # Renaming for joining
  mutate(Latitude = as.numeric(Latitude)) %>%
  mutate(Longitude = as.numeric(Longitude))

rth_ebird <- rth_ebird %>%
  mutate(Latitude = as.numeric(Latitude)) %>%
  mutate(Longitude = as.numeric(Longitude))

rth_some <- rth_ebird %>% full_join(rth_inat)

jn_table <- jn_table %>%
  mutate(X = as.character(X))

rth_more <- rth_some %>% full_join(jn_table)

rth_most <- rth_some %>% full_join(rth_more)

all_rth <- rth_most %>% full_join(npn_rth)

all_rth <- all_rth[!duplicated(as.list(all_rth))]
all_rth <- all_rth %>%
  distinct()

```

- Parsing and mutating data

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

all_rth <- all_rth %>%
  mutate(Date = parse_date_time(Date, orders = c("mdy", "ymd", "dmy"))) %>%
  mutate(year_day = yday(Date))

```

- Filtering Ruby-throat data based on their migration timeframe of February to May [9]
- Finding the early Ruby-throat arrival dates
- Bands are further filtered by month to account for their usual arrival times based on Courter et al., 2013 and Journey North maps

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

rth_early <- all_rth %>%
  filter(Year %in% c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013) & Month %in% c(2,3,4,5))

rthearly30 <- rth_early %>%
  filter(Month %in% c(2,3)) %>%
  filter(Latitude > 29 & Latitude < 33) %>%
  mutate(mean = mean(year_day, na.rm = TRUE))

rthearly33 <- rth_early %>%
  filter(Month %in% c(3,4)) %>%
  filter(Latitude > 33 & Latitude < 36) %>%
  mutate(mean = mean(year_day, na.rm = TRUE))

rthearly36 <- rth_early %>%
  filter(Month %in% c(3,4)) %>%
  filter(Latitude > 36 & Latitude < 39) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rthearly39 <- rth_early %>%
  filter(Month %in% c(4,5)) %>%
  filter(Latitude > 39 & Latitude < 42) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rthearly42 <- rth_early %>%
  filter(Month %in% c(4,5)) %>%
  filter(Latitude > 42 & Latitude < 45) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) 

rth_early_all <- rbind(rthearly30, rthearly33, rthearly36, rthearly39, rthearly42)


```

- 2014-2025 Ruby-throat dataset:

```{r, warning=FALSE, message=FALSE}
#| code-fold: TRUE

rth_new <- all_rth %>%
  filter(Year %in% c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & Month %in% c(2, 3, 4, 5)) %>%
  mutate(mean0013 = 
  case_when(Latitude > 30 & Latitude < 33 ~ 77,
            Latitude > 33 & Latitude < 36 ~ 110.6667,
            Latitude > 36 & Latitude < 39 ~ 112.6667,
            Latitude > 39 & Latitude < 42 ~ 137.125,
            Latitude > 42 & Latitude < 45 ~ 141
                              ))                          
rth_30 <- rth_new %>%
  filter(Month %in% c(2,3)) %>%
  filter(Latitude > 30 & Latitude < 33) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean0013)

rth_33 <- rth_new %>%
  filter(Month %in% c(3,4)) %>%
  filter(Latitude > 33 & Latitude < 36) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean0013)

rth_36 <- rth_new %>%
  filter(Month %in% c(3,4)) %>%
  filter(Latitude > 36 & Latitude < 39) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean0013)

rth_39 <- rth_new %>%
  filter(Month %in% c(4,5)) %>%
  filter(Latitude > 39 & Latitude < 42) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean0013)

rth_42 <- rth_new %>%
  filter(Month %in% c(4,5)) %>%
  filter(Latitude > 42 & Latitude < 45) %>%
  mutate(mean = mean(year_day, na.rm = TRUE)) %>%
  mutate(difference = mean - mean0013)

rth_new_all <- rbind(rth_30, rth_33, rth_36, rth_39, rth_42)

```

## 8. Create Color Palettes

```{r}

#Mapping colors to "difference" columns

cpal <- colorNumeric(palette = "viridis", domain = fl_new_all$difference, na.color = NA)

rpal <- colorNumeric(palette = "viridis", domain = rth_new_all$difference, na.color = NA)
 
```

# Results

### Trends in Flower Blooms (2014 - 2025)

```{r}

leaflet(fl_new_all) %>%
  addProviderTiles(providers$Esri.WorldStreetMap, group = "World Map") %>%
  addProviderTiles(providers$OpenTopoMap, group = "Topo Map") %>%
  addLayersControl(
    baseGroups = c("World Map","Topo Map"),
    options = layersControlOptions(collapsed = FALSE)
    ) %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~longitude, ~ latitude, radius = 5, stroke = FALSE, color = ~cpal(difference), label = ~scientific_name) %>%
  addLegend(pal = cpal, values = ~difference, title = "Difference From Early Average (days)", position = "bottomleft") %>%
  setView(-94.5786, 39.0997, zoom = 4)

```

Dates for each latitude band:

-   30-33: April 3rd (About 11 days later)

-   33-36: April 3rd (About 11 days later)

-   36-39: April 4th (About 6 days earlier)

-   39-42: April 3rd (About 7 days earlier)

-   42-45: April 6th (About 19 days earlier)

-   45-50: April 4th (About 20 days earlier)

### Trends in Ruby-Throated Hummingbird Sightings (2014 - 2025)

```{r}

leaflet(rth_new_all) %>%
  addProviderTiles(providers$Esri.WorldStreetMap, group = "World Map") %>%
  addProviderTiles(providers$OpenTopoMap, group = "Topo Map") %>%
  addLayersControl(
    baseGroups = c("World Map","Topo Map"),
    options = layersControlOptions(collapsed = FALSE)
    ) %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~Longitude, ~ Latitude, radius = 5, stroke = FALSE, color = ~rpal(difference), label = ~Age.Sex) %>%
  addLegend(pal = rpal, values = ~difference, title = "Difference From Early Average (days)", position = "bottomleft") %>%
  setView(-94.5786, 39.0997, zoom = 4)

```

Dates for each latitude band:

-   30-33: March 14th (About 4 days earlier)

-   33-36: April 10th (About 11 days earlier)

-   36-39: April 21st (About 2 days earlier)

-   39-42: May 13th (About 4 days earlier)

-   42-45: May 17th (About 4 days earlier)

### Average Bloom/Arrival Times (2005-2013)

```{r}

early_plot <- ggplot() + geom_point(data = fl_early_all, aes(x = mean, y = latitude, color = mean)) + geom_smooth(data = fl_early_all, aes(x = mean, y = latitude), method = "lm", color = "darkred") + geom_point(data = rth_early_all, aes(x = mean, y = Latitude)) + geom_smooth(data = rth_early_all, aes(x = mean, y = Latitude, color = mean), method = "lm", color = "darkblue") + labs(title = "Average Bloom/Arrival Times (2005-2013)", x = "Mean Arrival/Flower Date (Year Days)", y = "Latitude") + theme(plot.title = element_text(family = "serif", size = 12), axis.title = element_text(family = "serif", size = 10), panel.background = element_blank(), panel.grid = element_line(color = "lightblue"))

ggplotly(early_plot)

```

Generally, Red Columbine and Red Buckeye flowering is spread over a large portion of Ruby-throat migration.

### Average Bloom/Arrival Times (2014-2025)

```{r}

late_plot <- ggplot() + geom_point(data = fl_new_all, aes(x = mean, y = latitude, color = mean)) + geom_smooth(data = fl_new_all, aes(x = mean, y = latitude), method = "lm", color = "darkred") + geom_point(data = rth_new_all, aes(x = mean, y = Latitude)) + geom_smooth(data = rth_new_all, aes(x = mean, y = Latitude, color = mean), method = "lm", color = "darkblue") + labs(title = "Average Bloom/Arrival Times (2014-2025)", x = "Mean Arrival/Flower Date (Year Days)", y = "Latitude") + theme(plot.title = element_text(family = "serif", size = 12), axis.title = element_text(family = "serif", size = 10), panel.background = element_blank(), panel.grid = element_line(color = "lightblue"))

ggplotly(late_plot)

```

Generally, Red Columbine and Red Buckeye flowering happens over a shorter time within Ruby-throat spring migration.

### Observation Locations of Flowers

```{r}

leaflet(all_flower)%>%
  addProviderTiles(provider = "Esri") %>%
  setView(-85.7664, 38.2469, zoom = 4) %>%
  addHeatmap()

```

# Conclusions

Trends for Red Columbine and Red Buckeye show that the timing of first observations have somewhat changed from the 2005-2013 average. At the highest latitudes, there is not much change in flowering times. But for latitudes between 30 and 36, flowers have been sighted both earlier and later than in previous years. Since these flowers are spring ephemerals, warmer winter temperatures, or more extreme heat in spring and summer, may cause this trend, as these flowers will not bloom under extreme heat, or until they have gone through a period of sufficient cooling [10].

Additionally, the average first observation date for Ruby-throated hummingbirds has been generally shifting earlier with increasing latitudes.

These trends do somewhat align, especially at higher latitudes. So, these food sources will be available when Ruby-throats migrate back to higher latitudes earlier. But, Ruby-throats may may miss the flowering of these plants at lower latitudes.

It is important to note that this dataset is relatively small, and data used to acquire average arrival/bloom dates for 2005 to 2013 was sparse. Additionally, since this data is collected from citizen science projects, the data may be more biased towards urban areas, and its distribution is not even. As shown in the heatmap for the flowers, observations do tend to be concentrated in urban areas.

# References

1. Gill, J. A., Alves, J. A., Sutherland, W. J., Appleton, G. F., Potts, P. M., & Gunnarsson, T. G. (2014). Why is timing of bird migration advancing when individuals are not? Proceedings of the Royal Society B: Biological Sciences, 281(1774), 20132161. <https://doi.org/10.1098/rspb.2013.2161>

2. Xie, Y., Thammavong, H. T., Turner, A. L., Turner, B. I., & Park, D. S. (2025). Shifting spring ephemeral pollination windows under climate change – a three-body problem. New Phytologist, 248(1), 354–369. <https://doi.org/10.1111/nph.70373>

3. Ruby-throated Hummingbird Overview, All About Birds, Cornell Lab of Ornithology. (n.d.). Retrieved November 19, 2025, from <https://www.allaboutbirds.org/guide/Ruby-throated_Hummingbird/overview>

4. Ruby-throated Hummingbird. (n.d.). Retrieved November 19, 2025, from <https://www.fs.usda.gov/wildflowers/pollinators/pollinator-of-the-month/ruby-throated_hummingbird.shtml>

5. Wild Columbine (Plants of Connecticut) · iNaturalist. (n.d.). iNaturalist. Retrieved November 19, 2025, from <https://www.inaturalist.org/guide_taxa/373315>

6. Red Buckeye (Aesculus pavia). (n.d.). iNaturalist. Retrieved November 19, 2025, from <https://www.inaturalist.org/taxa/81708-Aesculus-pavia>

7. Red Columbine. (2004). Retrieved November 19, 2025, from <https://plants.usda.gov/DocumentLibrary/plantguide/pdf/pg_aqca.pdf>

8. USDA Plants Database Plant Profile General. (n.d.). Retrieved November 19, 2025, from <https://plants.usda.gov/plant-profile/AEPA>

9. Courter, J. R., Johnson, R. J., Bridges, W. C., & Hubbard, K. G. (2013). Assessing migration of Ruby-throated Hummingbirds ( Archilochus colubris ) at broad spatial and temporal scales. The Auk, 130(1), 107–117. <https://doi.org/10.1525/auk.2012.12058>

10. Kudo, G., & Cooper, E. J. (2019). When spring ephemerals fail to meet pollinators: Mechanism of phenological mismatch and its impact on plant reproduction. Proceedings of the Royal Society B: Biological Sciences, 286(1904), 20190573. <https://doi.org/10.1098/rspb.2019.0573>

Thanks for checking out my web site!
