---
title: "Phenological Changes in Pollination Pals"
author: Aubrey Monaco
subtitle: Synchrony Shifts in the Ruby-throated Hummingbird and Preferred Flowers
date: today
date-format: long
editor_options:
  chunk_output_type: console
---

# Introduction

With warming temperatures due to anthropogenic climate change, migrating birds are starting to move back to their breeding grounds earlier each year [1]. Along with this, flowers that certain migrating birds, like Ruby-throated Hummingbirds (*Archilochus colubris*), feed on may be changing their flowering times due to climate change [2].

![[Ruby-throated Hummingbird [3]]{style="color:darkblue;"}](data/rth_1.jpg){fig-align="center" width="400"}

I want to investigate how the Ruby-throated Hummingbird has been shifting its spring migration patterns at varying latitudes, and how the appearances of two spring ephemeral species it prefers to feed on during this migration, Red Buckeye (*Aesculus pavia*) and Red Columbine (*Aquilegia canadensis*), has changed (or not changed) over the past 10* years at varying latitudes [4].

*My maps and graphs are from 2014 to 2025 to account for any incomplete datasets from 2025

::: {layout-ncol=2}

![[Red Columbine [5]]{style="color:darkred;"}](data/cb_1.jpg){fig-align="center" width="200"}

![[Red Buckeye [6]]{style="color:darkred;"}](data/rb_1.jpg){fig-align="center" width="226"}

:::

I predict that Ruby-throated Hummingbirds are migrating back to their breeding grounds earlier each year, and that they are arriving at higher latitudes increasingly earlier than lower latitudes. Additionally, I suppose that the first sightings of these preferred flowers will be earlier at higher latitudes and later at lower latitudes compared to previous years. So, as Ruby-throated Hummingbirds migrate back earlier, these flowers may not be available at lower latitudes for them to feed on.

# Materials and Methods

:::{.panel-tabset}

## 1. 
### Install Packages and Load Libraries

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

#install.packages("leaflet") # Dynamic maps
#install.packages("leaflet.providers") # Different Basemaps
#install.packages("leaflet.extras") # Heatmaps
#install.packages("rinat") # Downloads iNaturalist data
#install.packages("rnpn") # Downloads US National Phenology Network Data
#install.packages("plotly") # Interactive graphs
library(doParallel) # Download data in a parallel loop
library(foreach) # Download data in a loop
library(plotly)
library(leaflet)
library(leaflet.providers)
library(leaflet.extras)
library(rinat)
library(rnpn)
library(knitr) # Making tables
library(rvest) # Used to download data from Journey North
library(dplyr) # Mutating and filtering data
library(lubridate) # Parsing date data
library(tidyr) # Cleaning data
library(ggplot2)# Plotting trends

```

## 2. 
### Flowers: Load iNaturalist and US NPN Data.

- foreach, doParallel, and rinat package iNaturalist data download: 

- taxon_name is set to each desired species, place_id is set to 1 for the United States, maxresults is set to 10000 to ensure all data is downloaded, and quality is set to "research" to ensure verified data

```{r message=FALSE, warning=FALSE, eval=FALSE}
#| code-fold: TRUE

#Registering cores for downloading data

num_cores <- parallel::detectCores() - 1
registerDoParallel(num_cores)

#Parallel data download for both species 2005-2019

nat_0519 <- foreach(s = c("Aesculus pavia", "Aquilegia canadensis"), .packages = c("dplyr", "rinat"), .errorhandling = "pass", .combine = bind_rows) %:% foreach(a = c(2005:2019), .packages = c("dplyr", "rinat"), .combine = bind_rows) %dopar% { 
print(paste("Downloading data for year:", a))
Sys.sleep(60)
get_inat_obs(taxon_name = s, year = a, place_id = 1, maxresults = 10000, quality = "research") 
}

#Sequential data download for both species 2020-2025

nat_2025 <- foreach(s = c("Aesculus pavia", "Aquilegia canadensis"), .packages = c("dplyr", "rinat"), .combine = rbind) %:% foreach(a = c(2020:2025), .packages = c("dplyr", "rinat"), .combine = rbind) %do% { 
print(paste("Downloading data for year:", a))
Sys.sleep(60)
get_inat_obs(taxon_name = s, year = a, place_id = 1, maxresults = 10000, quality = "research") 
}

```

- Individual datasets combined and written csv is loaded into environment as "nat_flower"

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

#fl_all <- rbind(nat_0519, nat_2025)

#write.csv(fl_all, "data/fl_all.csv")

nat_flower <- read.csv("data/fl_all.csv") 
nat_flower <- nat_flower %>%
  filter(common_name %in% c("red columbine", "Red Buckeye")) %>%
  mutate(year = year(observed_on))

```

- rnpn package for NPN data download (automatically saves as a csv): 

```{r, message=FALSE, warning=FALSE, eval=FALSE}
#| code-fold: TRUE

# Species IDs are available through the NPN's website (Red Buckeye = 784, Red Columbine = 147)

npn_download_individual_phenometrics(
 request_source='Aubrey Monaco', 
 years=c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
 species_id=c(784, 147), 
 download_path="data/npn_flower.csv"
 )

```

- NPN data loaded into environment

```{r}
#| code-fold: TRUE

npn_flower <- read.csv("data/npn_flower.csv") 
npn_flower <- npn_flower %>%
  rename(year = first_yes_year)

```

## 3.  
### Flowers: Combine the iNaturalist and NPN Data

```{r, message=FALSE}

# iNaturalist and NPN data are joined and duplicate observations are removed, leaving about 52,000 data points

all_flower <- nat_flower %>% full_join(npn_flower)
all_flower <- all_flower[!duplicated(as.list(all_flower))]
all_flower <- all_flower %>%
  distinct()

```

## 4. 
### Flowers: Standardize and Mutate Data

```{r, message=FALSE, warning=FALSE}

# dplyr and lubridate used to standardize and add columns from dates

all_flower <- all_flower %>%
  mutate(observed_on = parse_date_time(observed_on, orders = c("mdy", "ydm", "dmy"))) %>% # Standardizes date formats
  mutate(year_day = yday(observed_on)) %>% # Puts date into day out of 365
  mutate(month = month(observed_on)) # Pulls month from date
 
```

## 5. 
### Flowers: Calculate Means of Early Period (2005-2013)

- Found the mean date of first flower sightings from 2005 to 2013 and from March to May (when these flowers' bloom times overlap) [7, 8]
- Separated sightings into 3 bands of latitude to account for difference in average flowering times across their range (grouped by latitudes with similar "year_day" averages)

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

fl_early <- all_flower %>%
  filter(year %in% c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013) & month %in% c(3, 4, 5))

fl_early_all <- fl_early %>%
  mutate(latbin=cut(latitude, breaks = c(30,36,42,48), labels = c("30-36", "36-42", "42-48"))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE)) 
  
fl_means <- fl_early_all %>%
  group_by(latbin) %>%
  summarise(mean) %>%
  rename("Latitude" = "latbin", "Flowering Day Mean" = "mean")

#One row for each latitude band mean selected
selected_fl <- fl_means[c(1, 25, 39), ]

#Table to show early period flowering times based on latitude bands
kable(selected_fl, format = "pipe", caption = '<span style="color: darkred;">2005-2013 Flowering Time Means (days)</span>')

```

## 6. 
### Flowers: Calculate Means and Difference From 2014-2025 to 2005-2013

- Added a column (mean0513) based on "2005-2013 Flowering Time Means" table and assigned by corresponding latitudes
- Used 6 bands of latitude for more specific analysis
- Created a "difference" column, showing the difference in days of arrival between the early and current average using mean - mean0513
- Bound all of these bands into one dataset for plotting

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

fl_new <- all_flower %>%
  filter(year %in% c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & month %in% c(3, 4, 5)) %>%
  mutate(mean0513 = 
         case_when(
         latitude > 30 & latitude < 33 ~ 81.95652,
         latitude > 33 & latitude < 36 ~ 81.95652,
         latitude > 36 & latitude < 39 ~ 99.73333,
         latitude > 39 & latitude < 42 ~ 99.73333,
         latitude > 42 & latitude < 45 ~ 114.30769,
         latitude > 45 & latitude < 48 ~ 114.30769))

fl_new_all <- fl_new %>%
  mutate(latbin=cut(latitude, c(30,33,36,39,42,45,48), labels = c("30-33", "33-36", "36-39", "39-42","42-45", "45-48"))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE)) %>%
  mutate(difference = mean-mean0513) %>%
  drop_na(difference)

fl_new_difference <- fl_new_all %>%
  group_by(difference) %>%
  select(latbin, difference) %>%
  rename("Latitude" = "latbin", "Difference in Ruby-throat Arrival Dates" = "difference") 

#Selected 1 row for each latitude band difference
fl_new_difference <- fl_new_difference[c(1, 895, 222, 1425, 2656, 4115), ]
  
```

## 7. 
### Repeat Process for Ruby-throated Hummingbirds

- Used data from eBird and Journey North in addition to iNaturalist and the US NPN.

My eBird datasets can be found on figshare here:

-   1995-2021: <https://figshare.com/articles/dataset/eBird_rthhum_2010_2014/30535889?file=59303393>

-   2022-2025: <https://figshare.com/articles/dataset/eBird_rthhum_2015_2025/30535895?file=59303390>

Visit each link and click "download"

Or:

Download the Ruby-throated Hummingbird data from eBird by visiting: <https://media.ebird.org/catalog?birdOnly=true&taxonCode=rthhum>, filter by year (I did this in two separate actions \[1995-2021 and 2022-2025\] to ensure I could download all of the data), filter by location (United States), and download as a csv.

- iNaturalist data download:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
#| code-fold: TRUE

#Parallel data download for Ruby-throats 2005-2019

rth_0519 <- foreach(e = c(2005:2019), .packages = c("dplyr", "rinat"), .combine = bind_rows) %dopar% {
print(paste("Downloading data for year:", e))
Sys.sleep(45)
get_inat_obs(taxon_name = "Archilochus colubris", year = e, place_id = 1, maxresults = 10000, quality = "research") 
}

#Sequential data download for Ruby-throats 2020-2025

rth_2025 <- foreach(e = c(2020:2025), .packages = c("dplyr", "rinat"), .combine = bind_rows) %do% {
print(paste("Downloading data for year:", e))
Sys.sleep(45)
get_inat_obs(taxon_name = "Archilochus colubris", year = e, place_id = 1, maxresults = 10000, quality = "research") 
}

```

```{r, eval=FALSE}
#| code-fold: TRUE

#Combined all iNat data into one data set for Ruby-Throats

rth_nat <- rbind(rth_0519, rth_2025)

write.csv(rth_nat, "data/rth_nat.csv")

```

- NPN data download:

```{r, warning=FALSE, message=FALSE, eval=FALSE}
#| code-fold: TRUE

#Ruby-Throated Hummingbird = 352 (Species ID)

npn_download_individual_phenometrics(
 request_source='Aubrey Monaco', 
 years=c(2005, 2006, 2007, 2008, 2009, 2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024, 2025),
 species_id=c(352), 
 download_path="data/npn_rth.csv")

```

- NPN data loaded into environment:

```{r}
#| code-fold: TRUE

npn_rth <- read.csv("data/npn_rth.csv")
  npn_rth <- npn_rth %>%
  rename(Year = first_yes_year)

```

Journey North Data Download:

```{r, collapse=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
#| code-fold: TRUE

#Links for all years of data download (can only do one year at a time)

jn_05 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2005&submit=View+Data"
jn_06 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2006&submit=View+Data"
jn_07 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2007&submit=View+Data"
jn_08 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2008&submit=View+Data"
jn_09 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2009&submit=View+Data"
jn_10 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2010&submit=View+Data"
jn_11 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2011&submit=View+Data"
jn_12 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2005&submit=View+Data"
jn_13 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2013&submit=View+Data"
jn_14 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2014&submit=View+Data"
jn_15 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2015&submit=View+Data"
jn_16 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2016&submit=View+Data"
jn_17 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2017&submit=View+Data"
jn_18 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2018&submit=View+Data"
jn_19 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2019&submit=View+Data"
jn_20 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2020&submit=View+Data"
jn_21 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2021&submit=View+Data"
jn_22 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2022&submit=View+Data"
jn_23 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2023&submit=View+Data"
jn_24 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2024&submit=View+Data"
jn_25 <- "https://journeynorth.org/sightings/querylist.html?season=spring&map=hummingbird-ruby-throated-first&year=2025&submit=View+Data"

#Loop to download data for all links

jn_table <- foreach(urls = c(jn_05, jn_06, jn_07, jn_08, jn_09, jn_10, jn_11, jn_12, jn_13, jn_14, jn_15, jn_16, jn_17, jn_18, jn_19, jn_20, jn_21, jn_22, jn_23, jn_24, jn_25), .packages = c("dplyr", "rvest"), .combine = bind_rows) %dopar% {
Sys.sleep(30)
read_html(urls) %>%
html_elements("table") %>%
html_table() %>%
as.data.frame()
}

```

- Joining all data:

```{r, warning=FALSE, message=FALSE}
#| code-fold: TRUE

#write.csv(jn_table, "data/jn_table.csv")
jn_table <- read.csv("data/jn_table.csv") %>%
  rename(latitude = Latitude, longitude = Longitude) %>%
  mutate(X = as.numeric(X))

#Loaded, joined and standardized eBird data
rth_0513 <- read.csv("data/RTH_1995_2021.csv")
rth_1425 <- read.csv("data/RTH_2022_2025.csv")
rth_ebird <- rth_0513 %>% full_join(rth_1425)
rth_ebird <- rth_ebird %>%
  rename(latitude = Latitude, longitude = Longitude) %>%
  mutate(latitude = as.numeric(latitude)) %>%
  mutate(longitude = as.numeric(longitude))

#iNaturalist data loaded into environment and standardized
rth_inat <- read.csv("data/rth_nat.csv")
rth_inat <- rth_inat %>%
  mutate(latitude = as.numeric(latitude)) %>%
  mutate(longitude = as.numeric(longitude)) %>%
  rename(Date = observed_on)

#All data joined
rth_some <- rth_ebird %>% full_join(rth_inat)
rth_more <- rth_some %>% full_join(jn_table)
rth_most <- rth_some %>% full_join(rth_more)
all_rth <- rth_most %>% full_join(npn_rth)

#Removing duplicate rows and columns, leaving about 126,000 data points
all_rth <- all_rth[!duplicated(as.list(all_rth))]
all_rth <- all_rth %>%
  distinct()

```

- Parsing and mutating data:

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

all_rth <- all_rth %>%
  mutate(Date = parse_date_time(Date, orders = c("mdy", "ydm", "dmy"))) %>%
  mutate(year_day = yday(Date)) %>%
  mutate(Month = month(Date)) %>%
  mutate(Year = year(Date))

```

- Filtering to account for Ruby-throat migration arrival times for varying latitudes between February to May based on Journey North and Courter et al., 2013

```{r, message=FALSE, warning=FALSE}
#| code-fold: TRUE

rth_early <- all_rth %>%
  filter(Year %in% c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013) & Month %in% c(2,3,4,5))

rth_early_all <- rth_early %>%
  mutate(latbin=cut(latitude, breaks = c(30,33,36,39,42,45,48), labels = c("30-33","33-36", "36-39", "39-42", "42-45", "45-48"))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE))

rth_means <- rth_early_all %>%
  group_by(mean) %>%
  select(latbin, mean) %>%
  rename("Latitude" = "latbin", "Ruby-throat Average Arrival Dates (days)" = "mean")

#One row for each latitude band mean selected
selected_rth <- rth_means[c(5, 15099, 6639, 7940, 7395, 7), ]

kable(selected_rth, format = "pipe", caption = '<span style="color: darkblue;">2005-2013 Ruby-throat Arrival Means</span>')

```

- "mean0513" column values taken from "Ruby-throat Arrival Means" table and assigned based on corresponding latitudes
- 2014-2025 Ruby-throat dataset:

```{r, warning=FALSE, message=FALSE}
#| code-fold: TRUE

rth_new <- all_rth %>%
  filter(Year %in% c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025) & Month %in% c(2,3,4,5)) %>%
  mutate(mean0513 = 
  case_when(latitude > 30 & latitude < 33 ~ 79.80628,
            latitude > 33 & latitude < 36 ~ 93.53428,
            latitude > 36 & latitude < 39 ~ 105.52074,
            latitude > 39 & latitude < 42 ~ 117.30150,
            latitude > 42 & latitude < 45 ~ 124.95684,
            latitude > 45 & latitude < 48 ~ 129.7148                 ))                  

rth_new_all <- rth_new %>%
  mutate(latbin=cut(latitude, c(30,33,36,39,42,45,48), labels = c("30-33", "33-36", "36-39", "39-42", "42-45", "45-48"))) %>%
  group_by(latbin) %>%
  mutate(mean = mean(year_day, na.rm=TRUE)) %>%
  mutate(difference = mean-mean0513) %>%
  drop_na(difference)
  
rth_new_difference <- rth_new_all %>%
  group_by(difference) %>%
  select(latbin, difference) %>%
  rename("Latitude" = "latbin", "Difference in Ruby-throat Arrival Dates" = "difference")

#One row for each latitude band difference selected
rth_new_difference <- rth_new_difference[c(1, 14105, 13263, 8261, 24450, 92), ]

```

## 8. 
### Create Color Palettes

```{r}
#| code-fold: TRUE

#Mapping colors to "difference" columns

cpal <- colorNumeric(palette = "viridis", domain = fl_new_all$difference, na.color = NA)

rpal <- colorNumeric(palette = "viridis", domain = rth_new_all$difference, na.color = NA)
 
```

:::

# Results

## Where Is This Data Coming From?
### Observation Locations of Flowers and Ruby-throats

:::: {.columns}

::: {.column width="48%"}

```{r, message=FALSE}
#| code-fold: TRUE
#| label: map-left

leaflet(all_flower)%>%
  addProviderTiles(provider = "Esri") %>%
  setView(-85.7664, 38.2469, zoom = 4) %>%
  addHeatmap()

```
Flower Observation Locations
:::

::: {.column width="4%"}

:::

::: {.column width="48%"}

```{r}
#| code-fold: TRUE
#| label: map-right

all_rth_map <- all_rth %>%
  filter(!is.na(latitude) & !is.na(longitude))

leaflet(all_rth_map)%>%
  addProviderTiles(provider = "Esri") %>%
  setView(-85.7664, 38.2469, zoom = 4) %>%
  addHeatmap(lng = ~longitude, lat = ~latitude)

```
Ruby-throat Observation Locations

:::

::::

:::{.panel-tabset}

### Trends in Flowering Times (2014 - 2025)

```{r}
#| code-fold: TRUE

leaflet(fl_new_all) %>%
  addProviderTiles(providers$Esri.WorldStreetMap, group = "World Map") %>%
  addProviderTiles(providers$OpenTopoMap, group = "Topo Map") %>%
  addLayersControl(
    baseGroups = c("World Map","Topo Map"),
    options = layersControlOptions(collapsed = FALSE)
    ) %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~longitude, ~ latitude, radius = 5, stroke = FALSE, color = ~cpal(difference), label = ~scientific_name) %>%
  addLegend(pal = cpal, values = ~difference, title = "Difference From Early Average (days)", position = "bottomleft") %>%
  setView(-94.5786, 39.0997, zoom = 4)

```

### Difference in Flowering Times (Days)

```{r}
#| code-fold: TRUE

kable(fl_new_difference, format = "pipe", caption = '<span style="color: darkred;">2014-2025 Difference in Days</span>')

```

:::

::: {.panel-tabset}

### Trends in Ruby-throat Sightings (2014 - 2025)

```{r}
#| code-fold: TRUE

leaflet(rth_new_all) %>%
  addProviderTiles(providers$Esri.WorldStreetMap, group = "World Map") %>%
  addProviderTiles(providers$OpenTopoMap, group = "Topo Map") %>%
  addLayersControl(
    baseGroups = c("World Map","Topo Map"),
    options = layersControlOptions(collapsed = FALSE)
    ) %>%
  addSimpleGraticule(interval = 3) %>%
  addCircleMarkers(~longitude, ~ latitude, radius = 5, stroke = FALSE, color = ~rpal(difference), label = ~Age.Sex) %>%
  addLegend(pal = rpal, values = ~difference, title = "Difference From Early Average (days)", position = "bottomleft") %>%
  setView(-94.5786, 39.0997, zoom = 4)

```

### Difference in Arrival Times (Days)

```{r}
#| code-fold: TRUE

kable(rth_new_difference, format = "pipe", caption = '<span style="color: darkblue;">2014-2025 Difference in Days</span>')

```

:::

## Average Flowering/Arrival Times (2005-2013)

```{r, message=FALSE}
#| code-fold: TRUE

early_plot <- ggplot() + geom_point(data = fl_early_all, aes(x = mean, y = latitude, color = "Mean"), size = 0.1) + geom_smooth(data = fl_early_all, aes(x = mean, y = latitude, color = "Flowers"), method = "lm") + geom_point(data = rth_early_all, aes(x = mean, y = latitude, color = "Mean"), size = 0.1) + geom_smooth(data = rth_early_all, aes(x = mean, y = latitude, color = "Ruby-throats"), method = "lm") + labs(title = "Average Flowering/Arrival Times (2005-2013)", x = "Mean Arrival/Flower Date (Year Days)", y = "Latitude") + theme(plot.title = element_text(family = "serif", size = 12), axis.title = element_text(family = "serif", size = 10), panel.background = element_blank(), panel.grid = element_line(color = "lightblue"), legend.text = element_text(family = "serif", size = 10), axis.text = element_text(family = "serif", size = 10)) + scale_color_manual(name = "Legend", breaks = c("Mean", "Flowers", "Ruby-throats"), values = c("Mean" = "black", "Flowers" = "darkred", "Ruby-throats" = "darkblue"))

ggplotly(early_plot)

```

Generally, Red Columbine and Red Buckeye flowering is spread over a large portion of Ruby-throat migration.

## Average Flowering/Arrival Times (2014-2025)

```{r, message=FALSE}
#| code-fold: TRUE

late_plot <- ggplot() + geom_point(data = fl_new_all, aes(x = mean, y = latitude, color = "Mean"), size = 0.1) + geom_smooth(data = fl_new_all, aes(x = mean, y = latitude, color = "Flowers"), method = "lm") + geom_point(data = rth_new_all, aes(x = mean, y = latitude, color ="Mean"), size = 0.1) + geom_smooth(data = rth_new_all, aes(x = mean, y = latitude, color = "Ruby-throats"), method = "lm") + labs(title = "Average Flowering/Arrival Times (2014-2025)", x = "Mean Arrival/Flower Date (Year Days)", y = "Latitude") + theme(plot.title = element_text(family = "serif", size = 12), axis.title = element_text(family = "serif", size = 10), panel.background = element_blank(), panel.grid = element_line(color = "lightblue"), legend.text = element_text(family = "serif", size = 10), axis.text = element_text(family = "serif", size = 10)) + scale_color_manual(name = "Legend", breaks = c("Mean", "Flowers", "Ruby-throats"), values = c("Mean" = "black", "Flowers" = "darkred", "Ruby-throats" = "darkblue"))

ggplotly(late_plot)

```

Generally, Red Columbine and Red Buckeye flowering happens over a shorter time within Ruby-throat spring migration.

# Conclusions

Trends for Red Columbine and Red Buckeye show that the timing of their first observations have changed from the 2005-2013 average. At the highest latitudes, flowering times are later than the early average. But for latitudes between 30 and 36, flowers have been sighted later than in previous years. Since these flowers are spring ephemerals, warmer winter temperatures, or more extreme heat in spring and summer, may cause this trend, as these flowers will not bloom under extreme heat, or until they have gone through a period of sufficient cooling [10].

Additionally, the average first observation date for Ruby-throated hummingbirds has been generally shifting earlier with increasing latitudes.

These trends do somewhat align, especially at higher latitudes. So, Red Columbine and Red Buckeye will be available when Ruby-throats migrate back to higher latitudes earlier. But, Ruby-throats may may miss the flowering of these plants at lower latitudes.

It is important to note that this dataset is relatively small, and data used to acquire average arrival/bloom dates for 2005 to 2013 was sparse. Additionally, since this data is collected from citizen science projects, the data distribution may not be even. As shown in the observation heatmaps, sightings do tend to be concentrated in urban areas.

In the future, more data could be integrated into this code to create a better understanding of these trends. Additionally, it is important to understand what other Ruby-throat food sources are being affected not only during spring migration, but during breeding season, fall migration, and overwintering periods as well. Understanding trends like these can give insight into how climate change is affecting biodiversity and hopefully help motivate change and showcase why this biodiversity is important to protect.

# References

1. Gill, J. A., Alves, J. A., Sutherland, W. J., Appleton, G. F., Potts, P. M., & Gunnarsson, T. G. (2014). Why is timing of bird migration advancing when individuals are not? Proceedings of the Royal Society B: Biological Sciences, 281(1774), 20132161. <https://doi.org/10.1098/rspb.2013.2161>

2. Xie, Y., Thammavong, H. T., Turner, A. L., Turner, B. I., & Park, D. S. (2025). Shifting spring ephemeral pollination windows under climate change – a three-body problem. New Phytologist, 248(1), 354–369. <https://doi.org/10.1111/nph.70373>

3. Ruby-throated Hummingbird Overview, All About Birds, Cornell Lab of Ornithology. (n.d.). Retrieved November 19, 2025, from <https://www.allaboutbirds.org/guide/Ruby-throated_Hummingbird/overview>

4. Ruby-throated Hummingbird. (n.d.). Retrieved November 19, 2025, from <https://www.fs.usda.gov/wildflowers/pollinators/pollinator-of-the-month/ruby-throated_hummingbird.shtml>

5. Wild Columbine (Plants of Connecticut) · iNaturalist. (n.d.). iNaturalist. Retrieved November 19, 2025, from <https://www.inaturalist.org/guide_taxa/373315>

6. Red Buckeye (Aesculus pavia). (n.d.). iNaturalist. Retrieved November 19, 2025, from <https://www.inaturalist.org/taxa/81708-Aesculus-pavia>

7. Red Columbine. (2004). Retrieved November 19, 2025, from <https://plants.usda.gov/DocumentLibrary/plantguide/pdf/pg_aqca.pdf>

8. USDA Plants Database Plant Profile General. (n.d.). Retrieved November 19, 2025, from <https://plants.usda.gov/plant-profile/AEPA>

9. Courter, J. R., Johnson, R. J., Bridges, W. C., & Hubbard, K. G. (2013). Assessing migration of Ruby-throated Hummingbirds (*Archilochus colubris*) at broad spatial and temporal scales. The Auk, 130(1), 107–117. <https://doi.org/10.1525/auk.2012.12058>

10. Kudo, G., & Cooper, E. J. (2019). When spring ephemerals fail to meet pollinators: Mechanism of phenological mismatch and its impact on plant reproduction. Proceedings of the Royal Society B: Biological Sciences, 286(1904), 20190573. <https://doi.org/10.1098/rspb.2019.0573>

Thanks for checking out my web site!
